Grappa: Scalable Computing on Mass-Market Hardware for Irregular, Low-Locality Applications

### 2012-07-22 doc/asplos13/method.tex

To explore the performance of the Grappa runtime we have implemented three algorithms: breadth first search, betweenness centrality, and unbalanced tree search.  These algorithms were implemented for the Cray XMT (our baseline) and for Grappa.  Performance results for Grappa were obtained on a 144 node cluster.  Individual nodes of this cluster contain the hardware depicted in Table~\ref{table:grappanode}.  Nodes are interconnected with both 10G Ethernet and Infiniband.  For all but startup and configuration, they are configured to communicate over the Infiniband network for these experiments.

The metric we use is algorithmic time, which means startup and loading of the data structure (from disk) is not included in the measurement.  Data is collected on real systems, which means minor variations in runtime exist from run to run.  The average of multiple runs are used, and where appropriate, confidence in the result is reported (error bars, standard deviation, etc).

One question that must be answered when making a comparison between
the Cray XMT and a typical HPC cluster is what is a fair comparison --
these systems are quite different.  Three options immediately come to
mind: equal number of processing cores, equal number of network
interfaces, and equal dollars.  \checkme{We discounted the last option fairly
quickly, because it isn't a lasting data-point and, as will be evident
later, would be hugely unfair to the XMT.}  The first option, cores,
has some merit, but Grappa is designed for applications that have no
locality in their computation.  This means almost all of their memory
accesses are remote.  Hence, the factor that limits their performance
is not processing, but communicating.  Hence, we have chosen the
middle option, network interfaces as the way to normalize across the
XMT and our cluster.  Each processor in the XMT system has its own
network interface to access shared memory.  In the HPC cluster we use,
each processor (which contains up to 32 cores), has a single Infiniband interface.  Hence, for our results we scale up XMT processors one for one with full system nodes.

\subsection{Systems}
For measurements, we run Grappa on a 144-node cluster of AMD
Interlagos processors. Nodes have 32-core (every pair share a floating-point
unit) 2.1-GHz processors, 64GB of memory, and \TODO{xxx 40Gb network card}. \TODO{cluster
    uses xxx Infiniband switch}. We configure the nodes to have 32 1-GB
hugepages to minimize TLB misses for the random access patterns we
expect from irregular applications. The results we present are for
this machine, but we have also run Grappa on our own 12-node Intel
Xeon Westmere cluster.

@@

\subsection{Applications}
\TODO{
    Traversal of an unbalanced, unpartitioned tree captures the foundations for implementing most forms of irregular parallelism. 
Recursion is found in every irregular divide and conquer algorithm.
Loops, even regular ones, must be dynamically scheduled to get good
load balance when run on a large, non-uniform system. Nested
parallelism involves fine-grain creation of dynamic amounts of work.
Dataflow..
}

\paragraph{Unbalanced tree search in-memory (UTS-Mem)}
Unbalanced Tree Search (UTS) is a benchmark
for evaluating the programmability and performance of systems for
parallel applications that require dynamic load balancing
\cite{Olivier:uts2006}. It involves traversing an unbalanced implicit
tree: at each vertex, its number of children is sampled from some
probability distribution, and this number of new nodes are added to a
work queue to be visited. While this benchmark captures irregular,
dynamic parallelism, we want to evaluate performance on algorithms
with irregular memory access patterns.

We augment UTS by using the existing traversal code to
create a large tree in memory, and then we traverse the in-memory
tree. We call this benchmark UTS-Mem, and the timed portion is this
traversal of the in-memory tree. This in-memory traversal is
unknowledgeable of the tree structure beforehand.

@@

The Grappa version of the in-memory tree search uses the
asynchronous parallel for loop over a visited vertex's children 
list. A larger threshold can be set to take advantage of this 
locality by processing more child pointers sequentially in a single task.

@@

\paragraph{Betweenness Centrality}
An important measure of the importance of particular vertices in a network is betweenness centrality (BC) \cite{freeman1979centrality}. By this measure, the "importance" of each vertex is computed by finding the fraction of shortest paths that pass through it, which can optionally be approximated by only computing shortest paths using a subset of the vertices as starting points. BC can be useful for understanding which vertices may have the greatest impact, so in social networks this could be the primary person linking two communities. Because it requires multiple breadth-first traversals across the entire graph and in reverse, on power-law degree graphs, this algorithm exercises random accesses rate, and load balancing. It also requires fine-grained synchronization on updates to vertex centrality values because multiple paths will update the same vertex. BC is a kernel in the DARPA High Performance Computing Systems (HPCS) Scalable Synthetic Compact Applications graph analysis (SSCA\#2) benchmark\cite{ssca2}. Performance for BC is also measured in TEPS, where the number of traversed edges is the total number of edges in the graph multiplied by the number of random starting vertices used in the approximate computation. We use the XMT implementation of BC implemented as part of the GraphCT~\cite{GraphCT} library and a comparable Grappa implementation, both of which use the parallel BC algorithm developed by Bader et al.~\cite{bader:bc}.

>>>>

To explore the performance of the Grappa runtime we have implemented three algorithms: breadth first search, betweenness centrality, and unbalanced tree search.  These algorithms were implemented for the Cray XMT (our baseline) and for Grappa.  Performance results for Grappa were obtained on a 144 node cluster.  Individual nodes of this cluster contain the hardware depicted in Table~\ref{table:grappanode}.  Nodes are interconnected with both 10G ethernet and infiniband.  For all but startup and configuration, they are configured to communicate over the infiniband network for these experiments.

The metric we use is algorithmic time, which means startup and loading of the data structure (from disk) is not included in the measurement.  Data is collected on real systems, which means minor variations in runtime exist from run to run.  The average of multiple runs are used, and where appropriate, confidence in the result is reported.

One question that must be answered when making a comparison between the
Cray XMT and a typical HPC cluster is what is a fair comparison -- these
systems are quite different.  Three options immediately come to mind:
equal number of processing cores, equal number of network interfaces,
and equal dollars.  We discounted the last option fairly quickly,
because it isn't a lasting data-point -- the cost of hardware shifts
over time, skewing the interpretation of results.  The first option,
cores, has some merit, but Grappa is designed for applications that have
no locality in their computation.  This means almost all of their memory
accesses are remote.  The factor that limits their performance is not
processing, but communicating.  Hence, we have chosen the middle option,
network interfaces as the way to normalize across the XMT and our
cluster.  Each processor in the XMT system has its own network interface
to access shared memory.  In the HPC cluster we use, each processor
(which contains up to 32 cores, although we only use \checkme{6} in our
experiments), has a single infiniband interface.  Hence, for our results
we scale up XMT processors one for one with full system nodes.

\subsection{Systems}

For measurements, we run Grappa on a 144-node cluster of AMD Interlagos
processors. Nodes have 32-core (every pair share a floating-point unit)
2.1-GHz processors, 64GB of memory, and \checkme{a 40Gb Super-terrific
Happy-Lucky} infiniband network card.   The cluster uses a
\checkme{McNugget} infiniband switch. We configure the nodes to have 32
1-GB hugepages to minimize TLB misses for the random access patterns we
expect from irregular applications.\comment{The results we present are for this
machine, but we have also run Grappa on our own 12-node Intel Xeon
Westmere cluster, which performs similarly.}

@@

\subsection{Applications}

We have used three benchmarks to explore the performance of Grappa:

\comment{
    Traversal of an unbalanced, unpartitioned tree captures the foundations for implementing most forms of irregular parallelism. 
Recursion is found in every irregular divide and conquer algorithm.
Loops, even regular ones, must be dynamically scheduled to get good
load balance when run on a large, non-uniform system. Nested
parallelism involves fine-grain creation of dynamic amounts of work.
Dataflow..
}

\paragraph{Unbalanced tree search in-memory (UTS-Mem)} Unbalanced Tree
Search (UTS) is a benchmark for evaluating the programmability and
performance of systems for parallel applications that require dynamic
load balancing \cite{Olivier:uts2006}. It involves traversing an
unbalanced implicit tree: at each vertex, its number of children is
sampled from some probability distribution, and this number of new nodes
are added to a work queue to be visited. While this benchmark captures
irregular, dynamic \emph{computation}, we actually want to evaluate
performance of algorithms with irregular \emph{memory} access patterns. 
Thus we augment UTS by using the existing traversal code to create a
large tree in memory, and then we traverse the in-memory tree. We call
this benchmark UTS-Mem, and the timed portion is this traversal of the
in-memory tree. This in-memory traversal has no knowledgeable of the
tree structure beforehand.

@@

The Grappa version of the in-memory tree search uses the
asynchronous parallel for loop over a visited vertex's children 
list.  \comment{WTF? -M
A larger threshold can be set to take advantage of this 
locality by processing more child pointers sequentially in a single task.}

@@

\paragraph{Betweenness Centrality}
An important measure of the importance of particular vertices in a network is betweenness centrality (BC) \cite{freeman1979centrality}. By this measure, the ``importance'' of each vertex is computed by finding the fraction of shortest paths that pass through it, which can optionally be approximated by only computing shortest paths using a subset of the vertices as starting points. BC can be useful for understanding which vertices may have the greatest impact, so in social networks this could be the primary person linking two communities. Because it requires multiple breadth-first traversals across the entire graph and in reverse, on power-law degree graphs, this algorithm exercises random accesses rate, and load balancing. It also requires fine-grained synchronization on updates to vertex centrality values because multiple paths will update the same vertex. BC is a kernel in the DARPA High Performance Computing Systems (HPCS) Scalable Synthetic Compact Applications graph analysis (SSCA\#2) benchmark\cite{ssca2}. Performance for BC is also measured in TEPS, where the number of traversed edges is the total number of edges in the graph multiplied by the number of random starting vertices used in the approximate computation. We use the XMT implementation of BC implemented as part of the GraphCT~\cite{GraphCT} library and a comparable Grappa implementation, both of which use the parallel BC algorithm developed by Bader et al.~\cite{bader:bc}.
